Anomaly detection with kernel density estimation on manifolds

### 2022-06-28 paper/paper.Rmd
Rewrite pushforward and pullback

For any point $p \in M$, let $\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and $\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curves passing through $p$, i.e. $\gamma_1(0) = \gamma_2(0) = p$. We say $\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart, $(U,\varphi)$, at $p$ such that 
$$
(\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
$$ 
A *tangent vector* to a manifold $M$ at point $p$, denoted as $v_p$, is any equivalent class of the differentiable curves initialized at $p$. The set of all tangent vectors at $p$ defines the *tangent space* of $M$ at $p$, denoted as $T_pM$. The tangent space is a vector space of dimension $d$, equal to the dimension of $M$, and it does not depend on the chart $\varphi$ locally at $p$. The collection of all tangent spaces defines the *tangent bundle*, $TM = \cup_{p \in M}T_pM$.

Tangent vectors can also be seen as the directional derivatives at $p$. For a given coordinate chart $\varphi=(x^1,\dots,x^d)$, the tangent vectors defining partial
derivatives are denoted as $\frac{\partial}{\partial x^1}(p),\dots,\frac{\partial}{\partial x^d}(p)$, which defines a *basis* of the tangent space.
The tangent space $T_pM$ also admits a dual space $T^\star_pM$ called the *cotangent space* with corresponding *cotangent vectors*
$z_p: T^\star_pM \rightarrow \mathbb{R}^d$, and a basis denoted as $dx^1(p),\dots,dx^d(p)$.

@@

A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each point $p$ is a local inner product $T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is a $d\times d$ symmetric positive definite matrix and varies smoothly at $p$. Generally, we
omit the subscript $p$ and refer to $g$ as the Riemannian metric. The inner product between two vectors $u, v \in T_pM$ is written as $\langle u, v \rangle_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j}$, is assumed. A differentiable manifold $M$ endowed with the Riemannian metric $g$ on each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.

The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{\langle u,v \rangle_g}$ and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{\langle u,v \rangle_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$. It could also be used to define the line element $dl^2 = g_{ij}dx^i dx^j$ and the volume element $dV_g = \sqrt{\det(g)}dx^1 \dots dx^d$, where $(x^1,\dots,x^d)$ are the local coordinates of the chart $(U, \varphi)$. 
For a curve $\gamma: I \rightarrow M$, the length of the curve is
$$
l(\gamma) = \sqrt{\int_0^1 \|\gamma\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt} dt},
$$
where $\gamma(I) \subset U$. The volume of $W \subset U$ is defined as
$$
Vol(W) = \int_W \sqrt{\det(g)}dx^1 \dots dx^d,
$$
which is also called the *Riemannian measure* on $M$.

The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold. Intuitively, geodesics are the *straightest possible curves* in a Riemannian manifold [Section 7.2.3 of @Nakahara2018-zs].
A curve $\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$, the second-order ordinary differential equation is satisfied, 
$$
\frac{d^2 x^i}{dt^2} + \Gamma^i_{jk} \frac{d x^j}{dt} \frac{dx^k}{dt} = 0,
$$ 
where ${x^i}$ are the coordinates of the curve $\gamma$ and $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by 
$$
\Gamma^i_{jk} = \frac{1}{2} \sum_l g^{il} (\frac{\partial g_{il}}{\partial x^k} 
+ \frac{\partial g_{kl}}{\partial x^j} - \frac{\partial g_{jk}}{\partial x^l}).
$$ 
The geodesics have a constant speed with norm $\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc
length functional $l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}$ when the curves are defined over the interval $[0,1]$. 
The *geodesic distance* $d_g$ is the length of the shortest geodesic between two points on the manifold. For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the distance to the cut locus is the *injectivity radius* at $p \in M$. Therefore, the injectivity radius of the Riemannian manifold $(M,g)$, $\textit{inj}_gM$, is the infimum of the injectivity radii over all points on the manifold.

@@

Denote $B(p, r) \subset T_pM$ as an open ball centered at point $p$ with raidus $r$. Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$. Define $exp_p$ as the *exponential map* at point $p$, where $exp_p$ is a differentiable, bijective map of differentiable inverse (i.e. *diffeomorphism*) that maps a tangent vector $u \in B(0_p, r)$ to the endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying $\gamma(0)=p, \gamma^{-1}(0)=u$, and $\gamma(1)=exp_p(u)$. The exponential map moves point $p$ from $p$ at speed $u$ to an endpoint after covering the length of $\|u\|$ along the geodesic in one time unit. The inverse of the exponential map is called the *logarithm map*, $\log_p(q):= \exp^{-1}_p(q)$, which gives the vector from point $p$ to $q$. Also define the *geodesic ball* centered at $p$ of radius $r > 0$ as the image by the exponential map of $B(0_p, r) \subset T_pM$ with $r < \textit{inj}_gM$. Then we could interpolate a geodesic $\gamma$ between two points $p$ and $q$ with the exponential map and the logarithmic map, $\gamma(t) = \exp_p(t\log_p(q))$, and the geodesic distance is given by $d_g(p,q) = \|\log_p(q)\|_g$.

@@

Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors. Let $f$ be an embedding from the Riemannian manifold $(M,g)$ to another smooth manifold $E$. The pushforward $h=\varphi*g$ of the Riemannian metric $g$ along $\varphi \equiv f^{-1}$ is a linear map $f_*:TM \rightarrow TE$, which maps the tangent vectors $u_p$ at point $p \in M$ to the tangent vectors $f_*u_p$ at the mapping point $f(p) \in E$, and satisfies that for $u, v \in T_f(p)E$, 
$$
\langle u,v \rangle_{\varphi*g_p} = <df_p^{-1}u, df_p^{-1}v>_{g_p},
$$ 
where $df_p^{-1}$ is the Jacobian of $f^{-1}$. The tangent vectors $f_\star u_p$ are equivalent to the velocity vector of a curve $\gamma: I\rightarrow M$ passing through point $p$ at time zero with a constant speed $\gamma^{-1}(0)=u_p$. Similarly, the pullback maps the cotangent vectors $z_{f(p)}$ at $f(p) \in E$ to cotangent vectors at $p \in M$ acting on tangent vectors $u \in T_pM$.

@@

<!-- ## Canonical metric -->

<!-- For a compact Kahler manifold M, canonical metric can mean: (i) Kahler-Enstein metric or (ii) metric with constant scalar curvature which is the extremal metric of the Calabi functional. -->

@@

For any $u \in T_pM$, let $\gamma(u): t \rightarrow \exp(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$ and $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$. Define $\theta_p(q): T_pM \rightarrow \mathbb{R}$ the volume density function on $M$ as 
$$
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
$$ 
<!-- $$ -->
<!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
<!-- $$ --> 
Rewrite $\theta_p(u) = \theta_p(\log_p(u))$. The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, and $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates 
$$
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
$$

@@

<!-- ## Laplace-Beltrami operator -->

## Example: $S^1$

@@

The CDF of X is given by 
$$
P(X \leq x) = P(\cos(\theta) \leq x) \\
= P(\theta \geq \arccos(x)) \\
= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
= 1 - \frac{2}{\pi} \arccos(x).
$$ 
Then the density of X is the derivative of the CDF w.r.t. $x$ $$
f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
= \frac{2}{\pi \sqrt{1 - x^2}}.
$$ 
Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could
estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$
and compare it with its true density $f(x)$.

@@

We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by
<!-- \begin{equation} --> 
<!-- \label{eq:pelletier} --> 
$$
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
$$ 
<!-- \end{equation} --> 
where $d=1$, $h$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.

@@

Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by 
$$
\theta_{X_i}(p) = \sqrt{1 - X_i^2}.
$$ 
Therefore, the estimator for $p \in S^1$ is given by 
$$
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{1 - X_i^2}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
$$ 
where the bandwidth $r$ is selected by the plug-in bandwidth selector implemented by the R package *ks*.

>>>>

For any point $p \in M$, let $\gamma_1: (-\epsilon_1, \epsilon_1)\rightarrow M$ and $\gamma_2: (-\epsilon_2, \epsilon_2)\rightarrow M$ be two smooth curves passing through $p$, i.e. $\gamma_1(0) = \gamma_2(0) = p$. We say $\gamma_1$ and $\gamma_2$ are *equivalent* if and only if there exists a chart $(U,\varphi)$ at $p$ such that 
$$
(\varphi \circ \gamma_1)^\prime(0) = (\varphi \circ \gamma_2)^\prime(0).
$$ 
A *tangent vector* to a manifold $M$ at point $p$, denoted as $v_p$, is any equivalent class of the differentiable curves initialized at $p$. The set of all tangent vectors at $p$ defines the *tangent space* of $M$ at $p$, denoted as $T_pM$. The tangent space is a vector space of dimension $d$, equal to the dimension of $M$, and it does not depend on the chart $\varphi$ locally at $p$. The collection of all tangent spaces defines the *tangent bundle*, $TM = \cup_{p \in M}T_pM$.

Tangent vectors can also be seen as the directional derivatives at $p$. For a given coordinate chart $\varphi=(x^1,\dots,x^d)$, the tangent vectors defining partial
derivatives are denoted as $\frac{\partial}{\partial x^1}(p),\dots,\frac{\partial}{\partial x^d}(p)$, which defines a *basis* of the tangent space.
The tangent space $T_pM$ also admits a dual space $T^\star_pM$ called the *cotangent space* with the corresponding *cotangent vectors* $z_p: T^\star_pM \rightarrow \mathbb{R}^d$ and a basis denoted as $dx^1(p),\dots,dx^d(p)$.

@@

A Riemannian metric $g_p$ defined on the tangent space $T_pM$ at each point $p$ is a local inner product $T_pM \times T_pM \rightarrow \mathbb{R}$, where $g_p$ is a $d\times d$ symmetric positive definite matrix and varies smoothly at $p$. Generally, we
omit the subscript $p$ and refer to $g$ as the Riemannian metric. The inner product between two vectors $u, v \in T_pM$ is written as $\langle u, v \rangle_g = g_{ij}u^iv^j$ using the Einstein summation convention where implicit summation over all indices, $\sum_{i,j} g_{ij}u^iv^j$, is assumed. A differentiable manifold $M$ endowed with the Riemannian metric $g$ on each tangent space $T_pM$ is called a *Riemannian manifold* $(M,g)$.

The Riemannian metric $g$ can be used to define the norm of a vector $u$, $\|u\| = \sqrt{\langle u,v \rangle_g}$, and the angle between two vectors $u$ and $v$, $\cos\theta = \frac{\langle u,v \rangle_g}{\|u\| \|v\|}$, which are the geometric quantities induced by $g$. It could also be used to define the line element $dl^2 = g_{ij}dx^i dx^j$ and the volume element $dV_g = \sqrt{\det(g)}dx^1 \dots dx^d$, where $(x^1,\dots,x^d)$ are the local coordinates of the chart $(U, \varphi)$. 
For a curve $\gamma: I \rightarrow M$, the length of the curve is
$$
l(\gamma) = \sqrt{\int_0^1 \|\gamma^\prime(t)\|^2_g dt} = \sqrt{\int_0^1 g_{ij} \frac{dx^i}{dt} \frac{dx^j}{dt} dt},
$$
where $\gamma(I) \subset U$. The volume of $W \subset U$ is defined as
$$
Vol(W) = \int_W \sqrt{\det(g)}dx^1 \cdots dx^d,
$$
which is also called the *Riemannian measure* on $M$.

The *geodesics* of $M$ are the smooth curves that locally joins the points along the shortest path on the manifold. Intuitively, geodesics are the *straightest possible curves* in a Riemannian manifold [Section 7.2.3 of @Nakahara2018-zs].
A curve $\gamma: I \rightarrow M$ is a geodesic if for all indices $i,j,k$, the second-order ordinary differential equation is satisfied, 
$$
\frac{d^2 x^i}{dt^2} + \Gamma^i_{jk} \frac{d x^j}{dt} \frac{dx^k}{dt} = 0,
$$ 
where $\{x^i\}$ are the coordinates of the curve $\gamma$ and $\Gamma^i_{jk}$ is the *Christoffel symbol* defined by 
$$
\Gamma^i_{jk} = \frac{1}{2} \sum_l g^{il} (\frac{\partial g_{il}}{\partial x^k} 
+ \frac{\partial g_{kl}}{\partial x^j} - \frac{\partial g_{jk}}{\partial x^l}).
$$ 
The geodesics have a constant speed with norm $\| \gamma^\prime(t) \|$, and they are the local minimizers of the arc
length functional $l:\gamma \rightarrow \sqrt{\int_0^1 \| \gamma^\prime(t) \|_g^2 dt}$ when the curves are defined over the interval $[0,1]$. 
The *geodesic distance* $d_g$ is the length of the shortest geodesic between two points on the manifold. For a point $p \in M$, when the geodesic distance starting at $p$ is not minimized, we call such set of points the *cut locus* of $p$, and the distance to the cut locus is the *injectivity radius* at $p \in M$. Therefore, the injectivity radius of the Riemannian manifold $(M,g)$, $\textit{inj}_gM$, is the infimum of the injectivity radii over all points on the manifold.

@@

Denote $B(p, r) \subset T_pM$ as an open ball centered at point $p$ with raidus $r$. Then $B(0_p, r) = exp_p^{-1}(B(p,r))$ is an open neighborhood of $0_p$ in the tangent space at $p$, $T_pM$, where $exp_p$ is the *exponential map* at point $p$. The exponential map maps a tangent vector $u \in B(0_p, r)$ to the endpoint of the geodesic $\gamma: I \rightarrow M$ satisfying $\gamma(0)=p, \gamma^{\prime}(0)=u$, and $\gamma(1)=exp_p(u)$. It is a differentiable bijective map of differentiable inverse (i.e. *diffeomorphism*). Intuitively, the exponential map moves point $p$ to an endpoint  at speed $u$ after covering the length of $\|u\|$ along the geodesic in one time unit.

The inverse of the exponential map is called the *logarithm map*, denoted as $\log_p(q):= \exp^{-1}_p(q)$, which gives the tangent vector to get from point $p$ to $q$ in one unit time. Also define the *geodesic ball* centered at $p$ of radius $r > 0$ as the image by the exponential map of $B(0_p, r) \subset T_pM$ with $r < \textit{inj}_gM$. Then we could interpolate a geodesic $\gamma$ between two points $p$ and $q$ with the exponential map and the logarithmic map, $\gamma(t) = \exp_p(t\log_p(q))$, and the geodesic distance is given by $d_g(p,q) = \|\log_p(q)\|_g$. 
<!-- [TODO: reference in appendix]. -->

@@

Pushforward and pullback are two notions corresponding to the notions of tangent and cotangent vectors. 
Let $\phi: M \rightarrow E$ be a smooth map between the Riemannian manifold $(M,g)$ to another smooth manifold $E$. Then the differential of $\phi$ at point $p$ is a linear map $d\phi_p: T_pM \rightarrow T_{\phi (p)}E$, which pushes the tangent vector $u \in T_pM$ at point $p$ forward to the tangent vector $\phi_*u \in T_{\phi (p)}E$ at the mapping point $\phi(p)$.
The image of the tangent vector $u \in T_pM$ under the differential $d\phi_p$, denoted as $d\phi_p u$ is called the pushforward of $u$ by the map $\phi$.
Then pushforward metric $h=\varphi_*g$ of the Riemannian metric $g$ along $\varphi$ is given by the inner product
$$
\langle \phi_*u,\phi_*v \rangle_{\varphi*g} = \langle d\phi_p \phi_*u, d\phi_p \phi_*v \rangle_{g}.
$$

<!--  is a linear map $f_*:TM \rightarrow TE$, which maps the tangent vectors $u$ at point $p \in M$ to the tangent vectors $f_*u$ at the mapping point $f(p) \in E$.
For two tangent vectors $u, v \in T_{f(p)}E$, we have
$$
\langle u,v \rangle_{\varphi*g} = \langle df_p^{-1}u, df_p^{-1}v \rangle_{g},
$$ 
where $df_p^{-1}$ is the Jacobian of $f^{-1}$.  -->

The tangent vectors $\phi_*u$ are equivalent to the velocity vector of a curve $\gamma: I\rightarrow M$ passing through point $p$ at time zero with a constant speed $\gamma^{\prime}(0)=u$,
$$
d\phi_p(\gamma^{\prime}(0)) = (\phi \circ \gamma)^\prime (0).
$$
Similarly, the pullback maps the cotangent vectors $z_{f(p)}$ at $f(p) \in E$ to cotangent vectors at $p \in M$ acting on tangent vectors $u \in T_pM$. The linear map is called the pullback by $\phi$ and is often denoted as $\phi^*$.

@@

@@

For any tangent vector $u \in T_pM$, let $\gamma(u): t \rightarrow \exp_p(\frac{tu}{\|u\|})$ and let $w_i, i=1,\dots,d$ be Jacobi fields along $\gamma$ such that $w_i(0)=0$, for all $i=1,\dots,d$, $Dw_1/dt(0)=u/\|u\|$. Then $Dw_i/dt(0)$ forms an orthonomal basis of the tangent space $T_pM$. Define the volume density function on $M$, $\theta_p(q): T_pM \rightarrow \mathbb{R}$, as 
$$
\theta_p(u) = \|u\|^{d-1} \det(w_1(\|u\|),\dots,w_d(\|u\|)).
$$ 
<!-- $$ -->
<!-- \theta_p: q \rightarrow \theta_p(q)=\frac{\mu_{\exp_p{g}}}{\mu_{g_p}}(\exp^{-1}_p(q)), -->
<!-- $$ --> 
<!-- Rewrite $\theta_p(u) = \theta_p(\log_p(u))$.  -->
The exponential map $\exp_p: T_pM \rightarrow M$ induced by pullback of a volume form $\exp_p^*\textit{Vol}$ on $T_pM$, and $\theta_p$ is its density with respect to the Lebesgue measure of the Euclidean structure on $T_pM$, which gives in normal coordinates 
$$
d\exp_p^*\textit{Vo}l(u) = \theta_p(u)du.
$$
The volume density function is motivated by how a measure $du$ changes under the change of coordinates. It is related to the integration on manifolds and plays an important part in studing the density of manifolds.

@@

## Example of $S^1$ for manifold density estimation

@@

The CDF of X is given by 
$$
P(X \leq x) = P(\cos(\theta) \leq x) \\
= P(\theta \geq \arccos(x)) \\
= \int_{\arccos(x)}^{\pi/2} f(\theta) d\theta \\
=  \frac{2}{\pi} (\frac{\pi}{2} - \arccos(x) ) \\
= 1 - \frac{2}{\pi} \arccos(x).
$$ 
Then the density of X is the derivative of the CDF w.r.t. $x$ $$
f(x) = -\frac{2}{\pi} (-\frac{1}{\sqrt{1 - x^2}}) \\
= \frac{2}{\pi \sqrt{1 - x^2}}.
$$

Given $(x, y) \in \mathbb{R}^2$ as sample points from $S^1$, we could
estimate the density of each sample point $X_i \in S^1, i = 1, \dots, N$
and compare it with its true density $f(x)$.
We use the density estimator in @Pelletier2005-vu. For point $p \in S^1$, the density at $p$ is estimated by
<!-- \begin{equation} --> 
<!-- \label{eq:pelletier} --> 
$$
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r^d \theta_{X_i}(p)} K(\frac{d_g(p, X_i)}{r}),
$$ 
<!-- \end{equation} --> 
where $d=1$, $r$ is the bandwidth, $\theta_{X_i}(p)$ is the volume density function at sample point $X_i$, and $d_g(p, X_i)$ is the geodesic distance between $p$ and $X_i$ along the manifold $S^1$.

@@

@@

Then the volume density function $\theta_{X_i}(p)$ is the inner product of the tangent vectors produced by the logarithm map from $X_i$ to $p$, which is given by 
$$
\theta_{X_i}(p) = \sqrt{|\cos(d_g(p,q))|} = \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}.
$$ 
Therefore, the estimator for $p \in S^1$ is given by 
$$
f_N(p) = \frac{1}{N} \sum_{i=1}^{N} \frac{1}{r \sqrt{|\cos(\arccos(p) - \arccos(X_i))|}} K\bigg(\frac{\|\arccos(p) - \arccos(X_i))\|}{r}\bigg),
$$ 
where the bandwidth $r$ is selected with the plug-in bandwidth selector implemented by the R package *ks*.
